Sweat electrolytes and CFTR activity in the era of CFTR modulators

### 2024-01-31 Edith Zemanick/Sweat Test/sweat_test_analysis.Rmd
kmeans for sweat test analysis

```{r setup, include=FALSE}
library(DiagrammeR)
library(readxl)
library(arsenal)
library(tidyverse)
library(pROC)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if (Sys.info()["sysname"] == "Windows") {
  home_dir <- "C:/Users/timvigers/OneDrive - The University of Colorado Denver/Vigers/CF/Edith Zemanick/Sweat Test"
  source("C:/Users/timvigers/Documents/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Linux") {
  home_dir <- "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Darwin") {
  home_dir <- "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/genotype_class_severity.R")
}
knitr::opts_knit$set(root.dir = home_dir)
rm(home_dir)
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
n1 = nrow(df)
n2 = length(unique(df$PatientID))
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
# Diagnosis per Angela
df$Diagnosis <- factor(diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)])
levels(df$Diagnosis) = c("CFTR-RD or CRMS","CFTR-RD or CRMS","Cystic Fibrosis","Non-CF")
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars <- c("Amount", "Na", "K", "Cl")
num_vars <- c(paste0(num_vars, "_Left"), paste0(num_vars, "_Right"))
df[, num_vars] <- suppressWarnings(lapply(df[, num_vars], as.numeric))
# Filter per Maxie's code
exclude2 = c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)
no_dem = unique(df$PatientID[df$Demographics.Available!="TRUE" | is.na(df$Demographics.Available) | df$Sex==""])
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in% exclude2),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe", "Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF", "Severe_Unknown", "Unknown_Severe", "Unknown_Unknown"
  ),
  labels = c(
    "High", "Low", "Low", "Low", "Low", "Not CF", "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- cf_genotype_class_severity(df$Genotypes1, out = "class")
df$geno1_class[grep("508", df$Genotypes1)] <- "F508del"
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("I", "II", "III"), "MF")
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("IV", "V"), "RF")

@@

df$geno_class <- factor(paste0(
  pmin(df$geno1_class, df$geno2_class), "/",
  pmax(df$geno1_class, df$geno2_class)
))
# Which arm?
df$Arm <- apply(df[, c("Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left", "K_Right")], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` <- df$Na / df$Cl
high_ratio = unique(which(df$`Na/Cl Ratio`>2))
high_cl = unique(which(df$Cl>160))
high_na = unique(which(df$Na>160))
df = df[-unique(c(high_ratio,high_cl,high_na)),]
# Sweat chloride group
df$`Sweat Chloride Group` = cut(df$Cl,c(-Inf,30,60,Inf),right = F,
                                labels = c("< 30","30-59",">60"))
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols <- tableby.control(numeric.test = "kwt", numeric.stats = c("Nmiss", "median", "q1q3"))
labels <- list(
  Age.at.Test.in.Years = "Age (Years)",
  geno_risk = "Genotype Severity", geno_class = "Specific Genotype"
)
# First measure for everyone
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1) 
```

@@

@@

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ Sex + Age.at.Test.in.Years + Method+ fe(geno_risk) + fe(geno_class),
  data = t1_df, control = mycontrols, simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(
  `Sweat Chloride Group` ~ fe(geno_risk) + fe(Diagnosis)+  Cl + Na + `Na/Cl Ratio`,
  data = t1_df, control = mycontrols,simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r}

```

@@

For all participants with more than one sweat test, we compared their first and last values.

```{r results='asis'}
# Get first and last values for everyone with more than 1
multiples <- df %>%
  group_by(PatientID) %>%
  filter(n() > 1) %>% # Multiple values only
  filter(row_number() == 1 | row_number() == n()) %>% # First and last
  mutate(time = row_number())
multiples$time <- factor(multiples$time, levels = 1:2, labels = c("First", "Last"))
# Paired testing options
mycontrols <- tableby.control(
  numeric.test = "signed.rank", cat.test = "mcnemar",
  numeric.stats = c("Nmiss", "median", "q1q3")
)
```

```{r results='asis'}
p <- paired(
  time ~ Cl + Na + `Na/Cl Ratio`,
  signed.rank.exact = F, data = multiples,
  id = PatientID, control = mycontrols
)
summary(p, pfootnote = T, labelTranslations = labels)
```

@@

```{r}
roc_cf <- roc(response= t1_df$`Diagnosis`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("CFTR-RD or CRMS","Cystic Fibrosis"),
              quiet = T)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(response= t1_df$`Diagnosis`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("Non-CF","CFTR-RD or CRMS"),
              quiet = T)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)

@@

```{r warning=FALSE}
ggplot(t1_df[!is.na(t1_df$`Diagnosis`), ], 
       aes(log(`Na/Cl Ratio`), fill = `Diagnosis`)) +
  geom_histogram(alpha = 0.3, position = "identity",bins = 50) +
  geom_vline(xintercept = log(coords_cf$threshold), color = "blue") +
  geom_vline(xintercept = log(coords_non_cf$threshold), color = "green") +
  ylab("Count") + xlab("ln(Na/Cl Ratio)") +
  theme_bw()
```

@@

```{r}
# Re-format diagnosis column
t1_df$`Genotype Risk` <- factor(t1_df$geno_risk,
                     levels = c("Not CF","Low","High","Unknown"),
                     ordered = T)
t1_df$`Genotype Risk`[t1_df$`Genotype Risk`=="Unknown"]= NA
```

```{r}
roc_cf <- roc(response= t1_df$`Genotype Risk`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("Low","High"),
              quiet = T)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(response= t1_df$`Genotype Risk`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("Not CF","Low"),
              quiet = T)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)

@@

```{r warning=FALSE}
ggplot(t1_df[!is.na(t1_df$`Genotype Risk`), ], 
       aes(log(`Na/Cl Ratio`), fill = `Genotype Risk`)) +
  geom_histogram(alpha = 0.3, position = "identity",bins = 50) +
  geom_vline(xintercept = log(coords_cf$threshold), color = "blue") +
  geom_vline(xintercept = log(coords_non_cf$threshold), color = "green") +
  ylab("Count") + xlab("ln(Na/Cl Ratio)") +
  theme_bw()
```

@@

```{r}
# Exclude those with Cl >= 60
t1_df = t1_df %>% filter(Cl < 60)
```

```{r}
roc_cf <- roc(response= t1_df$`Diagnosis`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("CFTR-RD or CRMS","Cystic Fibrosis"),
              quiet = T)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(response= t1_df$`Diagnosis`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("Non-CF","CFTR-RD or CRMS"),
              quiet = T)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)

@@

```{r}
roc_cf <- roc(response= t1_df$`Genotype Risk`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("Low","High"),
              quiet = T)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(response= t1_df$`Genotype Risk`,
              predictor = t1_df$`Na/Cl Ratio`, 
              levels = c("Not CF","Low"),
              quiet = T)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)

>>>>

```{r setup, include=FALSE}
library(DiagrammeR)
library(readxl)
library(arsenal)
library(tidyverse)
library(factoextra)
library(pROC)
library(knitr)
source("/Users/timvigers/GitHub/CF/genotype_class_severity.R")
switch(Sys.info()[["sysname"]],
  Windows = {
    home_dir <- "C:/Users/timvigers/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Test"
  },
  Darwin = {
    home_dir <- "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Test"
  }
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
n1 <- nrow(df)
n2 <- length(unique(df$PatientID))
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
# Diagnosis per Angela
df$Diagnosis <- factor(diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)])
levels(df$Diagnosis) <- c("CFTR-RD or CRMS", "CFTR-RD or CRMS", "Cystic Fibrosis", "Non-CF")
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars <- c("Amount", "Na", "K", "Cl")
num_vars <- c(paste0(num_vars, "_Left"), paste0(num_vars, "_Right"))
df[, num_vars] <- suppressWarnings(lapply(df[, num_vars], as.numeric))
# Filter per Maxie's code
exclude2 <- c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)
no_dem <- unique(df$PatientID[df$Demographics.Available != "TRUE" | is.na(df$Demographics.Available) | df$Sex == ""])
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in% exclude2),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe", "Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF", "Severe_Unknown", "Unknown_Severe", "Unknown_Unknown"
  ),
  labels = c(
    "High", "Low", "Low", "Low", "Low", "Not CF", "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- cf_genotype_class_severity(df$Genotypes1, out = "class")
df$geno1_class[grep("508", df$Genotypes1)] <- "F508del"
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("I", "II", "III"), "MF")
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("IV", "V"), "RF")

@@

df$geno_class <- factor(paste0(
  pmin(df$geno1_class, df$geno2_class), "/",
  pmax(df$geno1_class, df$geno2_class)
))
# Which arm?
df$Arm <- apply(df[, c("Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left", "K_Right")], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` <- df$Na / df$Cl
high_ratio <- unique(which(df$`Na/Cl Ratio` > 2))
high_cl <- unique(which(df$Cl > 160))
high_na <- unique(which(df$Na > 160))
df <- df[-unique(c(high_ratio, high_cl, high_na)), ]
# Sweat chloride group
df$`Sweat Chloride Group` <- cut(df$Cl, c(-Inf, 30, 60, Inf),
  right = F,
  labels = c("< 30", "30-59", ">60")
)
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols <- tableby.control(
  numeric.test = "kwt",
  numeric.stats = c("Nmiss", "median", "q1q3")
)
labels <- list(
  Age.at.Test.in.Years = "Age (Years)",
  geno_risk = "Genotype Severity", geno_class = "Specific Genotype"
)
# First measure for everyone
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1)
# Order groups
df$Diagnosis <- factor(df$Diagnosis, levels = c("Cystic Fibrosis", "CFTR-RD or CRMS", "Non-CF"))
```

@@

# Cohorts by Na/Cl ratio

## Based on visual inspection

```{r}
# Visual inspection
ggplot(df, aes(x = Diagnosis, y = `Na/Cl Ratio`)) +
  geom_boxplot() +
  geom_hline(yintercept = 1, color = "red") +
  geom_hline(yintercept = 1.25, color = "red")
df$nacl_ratio_cat <- cut(df$`Na/Cl Ratio`, breaks = c(-Inf, 1, 1.25, Inf), right = F)
kable(table(df$nacl_ratio_cat, df$Diagnosis))
```

## Based on k-means clustering

```{r}
set.seed(1017)
# K means
km_df <- df %>%
  select(Diagnosis, Na, Cl) %>%
  drop_na()
ggplot(km_df, aes(x = Na, y = Cl, color = Diagnosis)) +
  geom_point()
km <- kmeans(km_df[, c("Na", "Cl")], centers = 3)
km_df$Cluster <- factor(km$cluster,
  levels = 1:3,
  labels = c("High", "Medium", "Low")
)
centers <- data.frame(km$centers)
centers$Cluster <- rownames(km$centers)
centers$Cluster <- factor(centers$Cluster,
  levels = 1:3,
  labels = c("High", "Medium", "Low")
)
ggplot(km_df, aes(x = Na, y = Cl, color = Cluster)) +
  geom_point(shape = 20, alpha = 0.2) +
  geom_point(data = centers, size = 5)
rownames(centers) <- NULL
centers <- centers %>% column_to_rownames("Cluster")
kable(centers, caption = "K-means centers", row.names = T)
kable(as.data.frame.matrix(table(km_df$Cluster, km_df$Diagnosis)), row.names = T)
```

@@

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ Sex + Age.at.Test.in.Years + Method + fe(geno_risk) + fe(geno_class),
  data = t1_df, control = mycontrols, simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(
  `Sweat Chloride Group` ~ fe(geno_risk) + fe(Diagnosis) + Cl + Na + `Na/Cl Ratio`,
  data = t1_df, control = mycontrols, simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

For all participants with more than one sweat test, we used only their first value to calculate the absolute difference (AD) between left and right arm.

```{r}
t1_df$`Cl AD` <- abs(t1_df$Cl_Left - t1_df$Cl_Right)
t1_df$`Na AD` <- abs(t1_df$Na_Left - t1_df$Na_Right)
```

### By diagnosis

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ `Cl AD` + `Na AD`,
  data = t1_df, control = mycontrols,
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

### By sweat chloride group

```{r results='asis'}
t1 <- tableby(
  `Sweat Chloride Group` ~ `Cl AD` + `Na AD`,
  data = t1_df, control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

### By genotype risk

```{r results='asis'}
t1 <- tableby(
  geno_risk ~ `Cl AD` + `Na AD`,
  data = t1_df, control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

For all participants with more than one sweat test, we compared their first and last values by calculating the absolute difference (AD) between first and last measure.

```{r results='asis'}
# Get first and last values for everyone with more than 1
multiples <- df %>%
  group_by(PatientID) %>%
  filter(n() > 1) %>% # Multiple values only
  filter(row_number() == 1 | row_number() == n()) %>% # First and last
  summarise(
    `Cl AD` = abs(last(Cl) - first(Cl)), # Calculate AD
    `Na AD` = abs(last(Na) - first(Na)),
    Diagnosis = first(Diagnosis),
    `Sweat Chloride Group` = first(`Sweat Chloride Group`),
    geno_risk = first(geno_risk)
  )
```

### By diagnosis

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ `Cl AD` + `Na AD`,
  data = multiples, control = mycontrols,
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

### By sweat chloride group

```{r results='asis'}
t1 <- tableby(
  `Sweat Chloride Group` ~ `Cl AD` + `Na AD`,
  data = multiples, control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

### By genotype risk

```{r results='asis'}
t1 <- tableby(
  geno_risk ~ `Cl AD` + `Na AD`,
  data = multiples, control = mycontrols
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r}
roc_cf <- roc(
  response = t1_df$`Diagnosis`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("CFTR-RD or CRMS", "Cystic Fibrosis"),
  quiet = T
)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(
  response = t1_df$`Diagnosis`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("Non-CF", "CFTR-RD or CRMS"),
  quiet = T
)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)

@@

```{r warning=FALSE}
ggplot(
  t1_df[!is.na(t1_df$`Diagnosis`), ],
  aes(log(`Na/Cl Ratio`), fill = `Diagnosis`)
) +
  geom_histogram(alpha = 0.3, position = "identity", bins = 50) +
  geom_vline(xintercept = log(coords_cf$threshold), color = "blue") +
  geom_vline(xintercept = log(coords_non_cf$threshold), color = "green") +
  ylab("Count") +
  xlab("ln(Na/Cl Ratio)") +
  theme_bw()
```

@@

```{r}
# Re-format diagnosis column
t1_df$`Genotype Risk` <- factor(t1_df$geno_risk,
  levels = c("Not CF", "Low", "High", "Unknown"),
  ordered = T
)
t1_df$`Genotype Risk`[t1_df$`Genotype Risk` == "Unknown"] <- NA
```

```{r}
roc_cf <- roc(
  response = t1_df$`Genotype Risk`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("Low", "High"),
  quiet = T
)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(
  response = t1_df$`Genotype Risk`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("Not CF", "Low"),
  quiet = T
)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)

@@

```{r warning=FALSE}
ggplot(
  t1_df[!is.na(t1_df$`Genotype Risk`), ],
  aes(log(`Na/Cl Ratio`), fill = `Genotype Risk`)
) +
  geom_histogram(alpha = 0.3, position = "identity", bins = 50) +
  geom_vline(xintercept = log(coords_cf$threshold), color = "blue") +
  geom_vline(xintercept = log(coords_non_cf$threshold), color = "green") +
  ylab("Count") +
  xlab("ln(Na/Cl Ratio)") +
  theme_bw()
```

@@

```{r}
# Exclude those with Cl >= 60
t1_df <- t1_df %>% filter(Cl < 60)
```

```{r}
roc_cf <- roc(
  response = t1_df$`Diagnosis`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("CFTR-RD or CRMS", "Cystic Fibrosis"),
  quiet = T
)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(
  response = t1_df$`Diagnosis`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("Non-CF", "CFTR-RD or CRMS"),
  quiet = T
)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)

@@

```{r}
roc_cf <- roc(
  response = t1_df$`Genotype Risk`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("Low", "High"),
  quiet = T
)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)

roc_non_cf <- roc(
  response = t1_df$`Genotype Risk`,
  predictor = t1_df$`Na/Cl Ratio`,
  levels = c("Not CF", "Low"),
  quiet = T
)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)
